---
title: "Lab 3: MAUP in Practice — Tessellations and Point-in-Polygon"
subtitle: "National Dam Inventory (NID): From Counts to Local Clustering"
format:
  html:
    code-fold: true
    toc: true
    include-before-body: ../slides/header.html
    include-after-body: ../slides/footer-annotations.html
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| include: false
knitr::opts_chunk$set(
  fig.width  = 8,
  message    = FALSE,
  warning    = FALSE,
  comment    = "",
  cache      = FALSE,
  fig.retina = 3
)
```

# Background

In this lab we will explore the impacts of tessellated surfaces and the
modifiable areal unit problem (MAUP) using the National Dam Inventory
maintained by the United States Army Corps of Engineers. The work
requires writing reusable functions and careful consideration of feature
aggregation/simplification, spatial joins, and data visualization. The
end goal is to visualize the distribution of dams and their purposes
across the country.

------------------------------------------------------------------------

---
title: "Lab 3: MAUP in Practice — Tessellations and Point-in-Polygon"
subtitle: "National Dam Inventory (NID): From Counts to Local Clustering"
author:
  - name: Maggie Rouin
    email: marguerite.rouin@colostate.edu
format: html
---

### Libraries

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false

# Data Science
library(tidyverse) # data wrangling
library(sf)        # Working with vector data
library(rmapshaper)# Simplify geometries
library(units)     # manage your units
library(mapview)
library(flextable)
library(janitor)
library(patchwork)
library(dplyr)

# Data
remotes::install_github("mikejohnson51/AOI") # get US counties
library(AOI) # county boundaries

# Visualization
library(gghighlight) # ggplot conditional highlighting
library(knitr) # table generation
library(kableExtra) # making tables pretty
```

------------------------------------------------------------------------

# Question 1:

Here we will prepare five tessellated surfaces from CONUS and write a
function to plot them in a descriptive way.

```{r}
us_counties <- aoi_get(state = "conus", county = "all") |> 
  st_transform(5070)
```

```{r}
us_centroid <- us_counties |> 
  st_centroid() 
```

------------------------------------------------------------------------

For this step:

-   Make a voroni tessellation over your county centroids (`MULTIPOINT`)
-   Make a triangulated tessellation over your county centroids
    (`MULTIPOINT`)
-   Make a gridded coverage with n = 70, over your counties object
-   Make a hexagonal coverage with n = 70, over your counties object

In addition to creating these 4 coverage's we need to add an ID to each
*tile*.

```{r}
voronoi <- us_centroid |> 
  st_combine() |> 
  st_cast("MULTIPOINT") |> 
  st_voronoi() |> 
  st_as_sf() |> 
  mutate(ID = 1:n())

triangulate <- us_centroid |> 
  st_combine() |> 
  st_cast("MULTIPOINT") |> 
  st_triangulate() |> 
  st_as_sf() |> 
  mutate(ID = 1:n())

gridded <-
  st_make_grid(us_counties, n = 70) |> 
  st_cast() |> 
  st_as_sf() |> 
  mutate(ID = 1:n())

hexagon <-
  st_make_grid(us_counties, n = 70, square = FALSE) |> 
  st_cast() |> 
  st_as_sf() |> 
  mutate(ID = 1:n())
```

### Step 1.4

If you plot the above tessellations you'll see that the **Voronoi**,
**Delaunay triangulation**, **hexagon**, and **square grids** all
produce regions far beyond the boundaries of CONUS.

```{r, eval = TRUE}
conus <- st_union(us_counties)

voronoi <- st_intersection(voronoi, conus)
triangulate <- st_intersection(triangulate, conus)
gridded <- st_filter(gridded, conus)
hexagon <- st_filter(hexagon, conus)
```

With a single feature boundary, we must carefully consider the
complexity of the geometry. Remember, the more points our geometry
contains, the more computations needed for spatial predicates our
differencing. For a task like ours, we do not need a finely resolved
coastal boarder.

```{r, eval = TRUE}
conus_simple <- conus |> 
  ms_simplify(, keep = 0.05)

ggplot(conus_simple) +
  geom_sf(fill = NA, color = "black", linewidth = 0.3) +
  theme_void()
```

::: callout-important
## Validation Checkpoint

After simplification, verify the boundary is still valid and covers all
dams:

```{r}
#| eval: false
# Check: Is the simplified boundary valid? TRUE
st_is_valid(conus_simple)

# Check: Does it still contain all of CONUS? NO (?)
st_bbox(conus_simple)  # Should cover roughly [-125, 24] to [-66, 49]

# Check: Visual comparison
plot(conus, main = "Original")
plot(conus_simple, main = "Simplified")
```
:::

```{r}
npts(conus)
#11292
npts(conus_simple)
#577

#I removed 10,715 points which I might discover is an oversimplification down the line. When comparing the plots, though, the two are very similar with simplification of the smaller islands off the east and west coasts. By removing this many points it is likely that this will impact data at the coasts since it is simplified.
```

```{r}
voronoi_conus <- st_intersection(voronoi, conus_simple)
triangulate_conus <- st_intersection(triangulate, conus_simple)

hexagon_conus <- st_filter(hexagon, conus_simple)
grid_conus <- st_filter(gridded, conus_simple)
```

```{r}
#| echo: true
#| eval: false

plot <- function(arg1, arg2) {
  ggplot(arg1)+
    geom_sf(fill = "white", color = "navy", linewidth = 0.2)+
    labs(caption = paste("n =", nrow(arg1), "features"))+
    theme_void()
}
```

Use your new function to plot each of your tessellated surfaces and the
original county data (5 plots in total):

```{r}
plot(voronoi_conus)
plot(triangulate_conus)
plot(grid_conus)
plot(hexagon_conus)
plot(us_counties)

#The plots vary in accuracy due to the simplification of the conus boundary, specifically with the Voronoi and Triangulation plots.
```

# Question 2:

In this question, we will write out a function to summarize our
tessellated surfaces.

```{r}
tesselation_summary <- function(arg1, arg2) {
  area_arg1 <- st_area(arg1) |> 
    set_units(km^2)
  
  area_num <- drop_units(area_arg1)
  
  summary_df <- data.frame(
    description = arg2,
    n_features = nrow(arg1),
    mean_area_km2 = mean(area_num, na.rm = TRUE),
    sd_area_km2 = sd(area_num, na.rm = TRUE),
    total_area_km2 = sum(area_num, na.rm = TRUE),
    stringsAsFactors = FALSE)
  
  return(summary_df)
}
```

Use your new function to summarize each of your tessellations and the
original counties.

```{r}
tesselation_summary(voronoi_conus, "Voronoi Tessellation")
tesselation_summary(triangulate_conus, "Delaunay Triangulation")
tesselation_summary(grid_conus, "Square Grid")
tesselation_summary(hexagon_conus, "Hexagonal Grid")
tesselation_summary(us_counties, "Original County Data")
```

Multiple `data.frame` objects can bound row-wise with `bind_rows` into a
single `data.frame`

```{r, eval = TRUE}
bound_summary <- bind_rows(
  tesselation_summary(triangulate_conus, "Delaunay Triangulation"),
  tesselation_summary(voronoi_conus, "Voronoi Tessellation"),
  tesselation_summary(hexagon_conus, "Hexagonal Grid"),
  tesselation_summary(grid_conus, "Square Grid"),
  tesselation_summary(us_counties, "Original County Data")
)
```

Once your 5 summaries are bound (2 tessellations, 2 coverage's, and the
raw counties) print the `data.frame` as a nice table using
`knitr`/`kableExtra`/`flextable` or your preferred table formatting
package.

```{r, echo = FALSE}
flextable(bound_summary) |> 
  set_header_labels(
    description = "Geometry Type",
    n_features = "Feature Count",
    mean_area_km2 = "Mean Area (km²)",
    sd_area_km2 = "SD Area",
    total_area_km2 = "Total Area (km²)") |>
  colformat_double(digits = 2, na_str = "N/A") |>
  autofit()
```

Comment on the traits of each tessellation. Be specific about how these
traits might impact the results of a point-in-polygon analysis in the
contexts of the modifiable areal unit problem and with respect
computational requirements.

The delaunay triangulation and voronoi tesselation both only have a
feature count of 1, whereas the hexagonal grid, square grid, and
original county data have higher feature counts but cover a smaller
total mean area. This may result in errors downstream since
point-in-polygon analysis allows you to understand how many points fall
within a certain polygon.

------------------------------------------------------------------------

# Question 3: Dam Distribution Analysis & Tessellation Selection

The data we will analyze in this lab are from the U.S. Army Corps of
Engineers National Dam Inventory (NID). This dataset documents \~91,000
dams in the United States and includes attributes such as design
specifications, risk level, age, and purpose.

Dam management agencies face a critical question: **Where are dams
actually clustered, and how do I visualize that clustering in a way that
supports infrastructure planning decisions?** The answer depends on your
choice of geographic boundaries—that is, your tessellation. Different
tessellations will reveal different patterns of dam concentration. Your
job in this question is to:

1.  Quantify dam distributions across multiple tessellation methods
2.  Understand how the **modifiable areal unit problem (MAUP)** creates
    different narratives from the same underlying data
3.  **Select and justify the tessellation that best supports management
    decisions** about dam clustering

For the remainder of this lab we will analyze the distributions of these
dams using point-in-polygon analysis.

In the tradition of this class - and true to data science/GIS work - you
need to find, download, and manage raw data. THe NID dataset is
available as a CSV download from the USACE website
[here](https://nid.sec.usace.army.mil/nid/#/downloads). Grab the CSV.

-   Return to your RStudio Project and read the data in using the
    `readr::read_csv`
    -   Use `janitor::clean_names()` to standardize column names to
        snake_case (tidy data convention)
    -   After reading the data in, be sure to remove rows that don't
        have location values (`!is.na()`)
    -   Convert the `data.frame` to a `sf` object by defining the
        coordinates and CRS
    -   Transform the data to a CONUS AEA (EPSG:5070) projection -
        matching your tessellation
    -   Filter to include only those within your CONUS boundary

```{r}
#| message: false
# Prefer a local NID CSV named `nation.csv`; otherwise download to that filename
nid_url <- "https://nid.sec.usace.army.mil/api/nation/csv"
dir.create("data", showWarnings = FALSE)
nid_dest <- "data/nation.csv"
if (file.exists(nid_dest)) {
  message("Using local NID file: ", nid_dest)
} else {
  tryCatch({
    message("Downloading NID CSV to: ", nid_dest)
    download.file(nid_url, nid_dest, quiet = FALSE, mode = "wb")
  }, error = function(e) {
    message("NID download failed: ", e$message)
  })
}

# read the CSV (will error if download failed and file missing)
if (!file.exists(nid_dest)) stop("NID CSV not found at ", nid_dest)
dams <- readr::read_csv(nid_dest, skip =1, show_col_types = FALSE) |> 
  janitor::clean_names()

usa <- AOI::aoi_get(state = "conus") %>% 
  st_union() %>% 
  st_transform(5070)

dams2 <- dams %>% 
  filter(!is.na(latitude)) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>% 
  st_transform(5070) %>% 
  st_filter(usa)
```

```{r}
pip_dams <- function(points, polygons, id_col) {
  st_geometry(points) <- "geometry"
  st_geometry(polygons) <- "geometry"
  
  if (st_crs(points) != st_crs(polygons)) {
  points <- st_transform(points, st_crs(polygons))}
  
  points_minimal <- points |> select(geometry)
  
  points_join <- st_join(points_minimal, polygons, join = st_intersects, 
                         left = FALSE) |> 
    st_drop_geometry() |> 
    count(.data[[id_col]], name = "n")
  
  polygons_join <- polygons |> 
    left_join(points_join, by = id_col) |> 
    mutate(n = replace_na(n, 0))
  
  return(polygons_join)
}
```

Apply your point-in-polygon function to each of your five tessellated
surfaces where:

```{r}
voronoi_pip <- pip_dams(dams2, voronoi_conus, id_col = "ID")
triangulate_pip <- pip_dams(dams2, triangulate_conus, id_col = "ID")
hexagon_pip <- pip_dams(dams2, hexagon_conus, id_col = "ID")
grid_pip <- pip_dams(dams2, grid_conus, id_col = "ID")
counties_pip <- pip_dams(dams2, us_counties, id_col = "fip_code")
```

### Step 3.4

Lets continue the trend of automating our repetitive tasks through
function creation. This time make a new function that extends your
previous plotting function.

```{r}
dam_plots <- function(obj, title){
  dam_n <- sum(obj$n, na.rm = TRUE)
  caption <- paste0("Number of dams: ", format(dam_n))
  
  ggplot(data = obj)+
    geom_sf(aes(fill = n))+
    scale_fill_viridis_c(option = "viridis", "Number of dams")+
    theme_void()
}
```

Apply your plotting function to each of the 5 tessellated surfaces with
Point-in-Polygon counts:

```{r}
voronoi_dam <- dam_plots(voronoi_pip, "Voronoi Tesselation of dams")
triangulate_dam <- dam_plots(triangulate_pip, "Triangulation Tesselation of dams")
hex_dam <- dam_plots(hexagon_pip, "Hexagonal Tessellation of dams")
grid_dam <- dam_plots(grid_pip, "Grid Tesselation of dams")
counties_dam <- dam_plots(counties_pip, "County-Level dams")
```

**Policy Analysis: Tessellation and Decision-Making**

Now that you have dam counts across all five tessellation methods,
compare the results. Specifically:

**a.** Create a side-by-side visualization (faceted or small multiples)
showing dam clustering under each of the five tessellations. Use
`patchwork` or `cowplot` to arrange the 5 plots you created in Step 3.5
for direct visual comparison.

```{r}
(voronoi_dam | triangulate_dam) / (hex_dam | grid_dam | counties_dam)
```

**b.** Identify **2–3 geographic regions** where tessellation choice
produces substantially *different* patterns (e.g., one method shows a
clear hotspot while another shows diffuse distribution). For each
region, describe the pattern difference and explain *why* the
tessellation geometry creates divergent results. Reference the
characteristics from your Q2 summary (number of tiles, mean area,
standard deviation).

The counties vs grid and hexagon maps show a difference in dam
distribution. The grid and hexagon maps indicate that a high density of
dams are found in middle america whereas the counties map indicates more
of a diffuse spread with additional hot spots on the east coast and near
Montana/Wyoming/Colorado. This is likely due to the fact that tiles are
weighted differently between those three maps which causes more of a
weight to a certain county than it would have if it were a square or
hexagon due to the area per tile.

**c.** **Which single tessellation would you recommend to a dam
operations manager seeking to understand geographic clustering?**
Justify your choice by addressing: - Does the tessellation align with
natural or management decision-making units (e.g., river basins, states,
hydrologic regions)? - Does it reveal meaningful patterns without
over-fragmenting the landscape or creating artificial boundaries? - How
does MAUP influence your recommendation—are other tessellations equally
valid, just for different questions? - How did simplification affect
your choice? Would a coarser or finer simplification (different `keep`
percentage) change your recommendation? - What does "best supports
management decisions" mean concretely in your context—does it matter
more that hot spots are obvious, or that boundaries align with
jurisdictions?

I would recommend a dam operations manger use the hexagon tessellation
map because it provides a clearer projection of the number of dams per
area covered. Since the hexagons are uniform, unlike the counties map,
it provides a better understanding of the areas where dams are
concentrated, which is highly aligned with the Mississippi River. Though
this map creates a simplification of the area covered by using the
hexagons compared to the grid map, which has smaller tiles, it provides
clearer boundaries.

The counties map provides easy boundary markers which could be
beneficial when considering workforce allocation, budgetary limitations,
or regulatory differneces by local governments since this can highly
impact management approaches. The main downfall, as discussed in the
previous question, is that it oversimplifies the scale used because
county size is not uniform which causes hot spots in areas that simply
have larger county sizes.

Your answer should be **3–4 paragraphs** and grounded in the actual
visualizations and patterns you observe from your tessellation
comparisons.

------------------------------------------------------------------------

# Question 4

The NID provides a comprehensive data dictionary
[here](https://files.hawaii.gov/dbedt/op/gis/data/nid_dams_data_dictionary.htm#Purposes).
In it we find that dam purposes are designated by a character
[code](https://files.hawaii.gov/dbedt/op/gis/data/nid_dams_data_dictionary.htm#Purposes).

These are shown below for convenience (built using knitr on a data.frame
called `nid_classifier`):

```{r}
#| echo: false
#| eval: true

nid_classifier = data.frame(matrix(c('I' , 'Irrigation',
'H' , 'Hydroelectric',
'C' , 'Flood Control',
'N' , 'Navigation',
'S' , 'Water Supply',
'R' , 'Recreation',
'P' , 'Fire Protection',
'F' , 'Fish and Wildlife',
'D' , 'Debris Control',
'T' , 'Tailings',
'G' , 'Grade Stabilization',
'O' , 'Other'), ncol = 2, byrow = T)) %>% 
  setNames(c("abbr", "purpose"))
```

```{r}
#| echo: false
#| eval: true

knitr::kable(nid_classifier,              
        caption = "NID 2019: Dam Purposes",
        format.args = list(big.mark = ",")) %>% 
    kableExtra::kable_styling("striped", full_width = TRUE, font_size = 14)
```

-   In the data dictionary, we see a dam can have *multiple* purposes.

-   In these cases, the purpose codes are concatenated in order of
    decreasing importance. For example, `SCR` would indicate the primary
    purposes are *Water Supply*, then *Flood Control*, then
    *Recreation.*

-   A standard summary indicates there are over 400 unique combinations
    of dam purposes:

```{r, echo = TRUE, eval = TRUE}
unique(dams2$purposes) %>% length()
#note: "PURPOSES" was used in the inital write up but purposes is correct and changed throughout
```

-   By storing dam codes as a concatenated string, there is no easy way
    to identify how many dams serve any one purpose... for example where
    are the hydro electric dams?
    -   No, the concatenated string data structure makes it difficult to
        view this information.

------------------------------------------------------------------------

To overcome this data structure limitation, we can identify how many
dams serve each purpose by splitting the PURPOSES values (`strsplit`)
and tabulating the unlisted results as a data.frame. Effectively this is
double/triple/quadruple counting dams bases on how many purposes they
serve:

```{r}
#| echo: false
#| eval: false
# create a vector of all characters in your purpose and unlist 
dam_freq <- strsplit(dams2$purposes, split = "") %>%
  unlist() %>% 
  table() %>% 
  as.data.frame() %>% 
  setNames(c("abbr", "count")) %>% 
  left_join(nid_classifier) %>% 
  mutate(lab = paste0(purpose, "\n(", abbr, ")"))
```

The result of this would indicate:

```{r}
#| echo: false
#| eval: false

ggplot() + 
  geom_col(data = dam_freq, aes(x = reorder(lab, -count), y = count, fill = purpose), alpha = .8) + 
  labs(x = "", y = "# Dams", title = "Number of Dams serving each purpose")  +
  scale_fill_viridis_d()+ 
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, vjust = .9, hjust = 1)) + 
  theme(legend.position = 'none')
```

-   Your task is to create point-in-polygon counts for at *least* 4 of
    the above dam purposes:

-   You will use `grepl` to filter the complete dataset to those with
    your chosen purpose

-   Remember that `grepl` returns a boolean if a given pattern is
    matched in a string

-   `grepl` is vectorized so can be used in `dplyr::filter`

------------------------------------------------------------------------

For your analysis, choose *at least* four of the above codes, and
describe why you chose them. Then for each of them, create a subset of
dams that serve that purpose using `dplyr::filter` and `grepl`

Finally, use your `point-in-polygon` function to count each subset
across your **cropped tessellation from Q1.4** (e.g., `hexagon_clipped`
or `square_clipped`).

```{r}
#chosen codes: R, F, S, and C since these were the codes with the highest number of dams associated

recreation_dams <- dams2 |> 
  filter(grepl("R", purposes))
fish_dams <- dams2 |> 
  filter(grepl("F", purposes))
supply_dams <- dams2 |> 
  filter(grepl("S", purposes))
flood_dams <- dams2 |> 
  filter(grepl("C", purposes))

recreation_pip <- pip_dams(recreation_dams, hexagon, id_col = "ID")
fish_pip <- pip_dams(fish_dams, hexagon, id_col = "ID")
supply_pip <- pip_dams(supply_dams, hexagon, id_col = "ID")
flood_pip <- pip_dams(flood_dams, hexagon, id_col = "ID")
```

-   Now use your plotting function from Q3 to map these counts on the
    **cropped tessellation**.

-   *But!* you will use `gghighlight` to **only** color those tiles
    where the count (*n*) is greater then the
    (`mean + 1 standard deviation`) of the set

-   Since your plotting function returns a `ggplot` object already, the
    `gghighlight` call can be added "`+`" directly to the function.

-   The result of this exploration is to highlight the areas of the
    country with the most

```{r}
p_recreaction <- dam_plots(recreation_pip) +
  gghighlight(n > (mean(n) + sd(n))) +
  labs(title = "Recreation")

p_fish <- dam_plots(fish_pip) +
  gghighlight(n > (mean(n) + sd(n))) +
  labs(title = "Fish and Wildlife")

p_supply <- dam_plots(supply_pip) +
  gghighlight(n > (mean(n) + sd(n))) +
  labs(title = "Water Supply")

p_flood <- dam_plots(flood_pip) +
  gghighlight(n > (mean(n) + sd(n))) +
  labs(title = "Flood Control")

(p_recreaction | p_fish) / (p_supply | p_flood)
```

### Step 4.3

Comment on the geographic distribution of dams you found. Does it make
sense? How might the tessellation you chose impact your findings? How
does the distribution of dams coincide with other geographic factors
such as river systems, climate, ect?

The dam distribution maps for fish and wildlife and water supply present
very similar spreads, with hot spots in the middle of the US from
roughly Montana through the midwest and down into Texas. This pattern
likely aligns with water needs for crops and would likely align with the
irrigation map if that variable was analyzed. The recreation map has
many hotspots along the eastern seaboard and through the midwest as
well, which is expected due to the high water availability for
recreation in those regions. Finally, the hot spots for the flood
control map are primarily in an area that aligns with the Mississippi
River.

# Question 5: Spatial Correlation of Dam Attributes — Introduction to Spatial Regression

Dam characteristics—such as **age, surface area, storage capacity, and
purpose**—vary geographically. Understanding the spatial structure of
these relationships is foundational to spatial regression modeling
(upcoming topics). In this question, you'll examine whether dam
attributes exhibit **spatial dependence** and assess whether neighboring
tiles have correlated characteristics.

**Aggregate Dam Attributes by Tile**

Using your **cropped tessellation from Q1.4** (your recommended grid
from Q3, clipped to CONUS boundary), compute **mean and variance** of
key dam attributes within each tile. The NID dataset typically includes:

-   **Age**: `year_completed` or similar (compute mean age per tile)
-   **Size**: `surface_area_acres` or `normal_pool_elevation` (compute
    mean per tile)
-   **Purpose codes**: `purposes` field (compute dominant purpose per
    tile; count of dams by purpose

**Produce visualizations** showing spatial patterns:

-   Map 1: Mean dam age by tile (use viridis scale)
-   Map 2: Mean surface area by tile
-   Map 3: Dam count by tile (you already have this from Q3)

Do you see geographic clustering of old vs. new dams? Large vs. small
dams?

```{r}
dam_aggregate <- dams2 |>
  st_join(hexagon_conus, join = st_intersects) |>
  st_drop_geometry() |>
  group_by(ID) |>
  summarise(n = n(),
    mean_year_completed = mean(year_completed, na.rm = TRUE),
    mean_surface_area   = mean(surface_area_acres, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(mean_age = 2026 - mean_year_completed) |>
  right_join(hexagon_conus, by = "ID") |>
  st_as_sf()

map1 <- ggplot(dam_aggregate) +
  geom_sf(aes(fill = mean_age), color = NA) +
  scale_fill_viridis_c(option = "viridis", name = "Years") +
  theme_void() +
  labs(title = "Mean Dam Age by Tile")

map2 <- ggplot(dam_aggregate) +
  geom_sf(aes(fill = mean_surface_area), color = NA) +
  scale_fill_viridis_c(option = "viridis", name = "Acres", 
                       trans = "log10") +
  theme_void() +
  labs(title = "Mean Surface Area by Tile (log scale)")

hex_dam <- dam_plots(hexagon_pip, "Hexagonal Tessellation of dams")

(map1)|(map2)|(hex_dam)
```

**Neighbor Correlation: Local Spatial Variation**

For your tessellation's neighbor matrix (from Q6), compute the **spatial
lag** of dam attributes: the average value of an attribute in a tile's
neighbors. This reveals whether *spatial neighbors have similar dam
attributes*—a key signal that observations are **not independent**
(violating OLS assumptions).

```{r}
#| eval: false
# Suggested workflow (write your own implementation):
# 1) build vectors for count, age, and size attributes by tile
# 2) handle NAs explicitly before matrix multiplication
# 3) compute spatial lag as W_std %*% attribute
# 4) correlate each attribute with its lag
# 5) report values and classify as weak/moderate/strong with your own thresholds
```

**Summarize in a table**:

```{r}
#| eval: false
# Create a compact summary table with:
# - attribute name
# - correlation with spatial lag
# - qualitative strength label
# Then add a 1-2 sentence interpretation below the table.
```

```{r}
standardize_weights <- function(geometry_sf) {

  # Check validity
  if (any(!st_is_valid(geometry_sf))) {
    geometry_sf <- st_make_valid(geometry_sf)
  }
  
  # Binary neighbor matrix
  W <- st_touches(geometry_sf, sparse = FALSE)
  diag(W) <- 0  # Remove self-neighbors
  
  # Row-standardize (handles isolated tiles)
  W_std <- W / rowSums(W)
  W_std[rowSums(W) == 0, ] <- 0  # Replace NaN (no neighbors) with 0
  
  return(W_std)
}

W_std <- standardize_weights(dam_aggregate)
W_std_county <- standardize_weights(counties_pip)

compute_spatial_lag <- function(attribute_vector, weight_matrix, 
                                original_na_mask = NULL) {
  if (is.null(original_na_mask)) {
    original_na_mask <- is.na(attribute_vector)
  }
  global_mean <- mean(attribute_vector, na.rm = TRUE)
  attribute_imputed <- ifelse(is.na(attribute_vector), 
                              global_mean, 
                              attribute_vector)  
  
spatial_lag <- weight_matrix %*% attribute_imputed
  spatial_lag <- as.vector(spatial_lag)
  
  return(list(
    lag = spatial_lag,
    imputed = attribute_imputed,
    na_mask = original_na_mask
  ))
}

# Dam count (n has no NAs — zeros for empty tiles, not NA)
lag_result_n    <- compute_spatial_lag(dam_aggregate$n, W_std)

# Mean year completed (NA where tile has no dams)
lag_result_year <- compute_spatial_lag(dam_aggregate$mean_year_completed, W_std)

# Mean surface area (NA where tile has no dams)
lag_result_area <- compute_spatial_lag(dam_aggregate$mean_surface_area, W_std)

# Correlation for each (use na_mask to exclude imputed values)
cor_count <- cor(lag_result_n$imputed, lag_result_n$lag, use = "complete.obs")
cor_age <- cor(lag_result_year$imputed[!lag_result_year$na_mask],
                lag_result_year$lag[!lag_result_year$na_mask], use = "complete.obs")
cor_size <- cor(lag_result_area$imputed[!lag_result_area$na_mask],
                lag_result_area$lag[!lag_result_area$na_mask], use = "complete.obs")

lag_result <- compute_spatial_lag(dam_aggregate$n, W_std)
lag_tess <- lag_result$lag
n_imputed <- lag_result$imputed
na_mask <- lag_result$na_mask

cor_value <- cor(lag_result$imputed[!lag_result$na_mask], lag_result$lag[!lag_result$na_mask], use = "complete.obs")

lag_county <- compute_spatial_lag(counties_pip$n, W_std_county)

#The correlation value is 0.824 which indicates that there is a strong correlation with neighboring tiles 
```

```{r}
classify_moran <- function(r) {
  abs_r <- abs(r)
  case_when(abs_r >= 0.5 ~ "Strong",
    abs_r >= 0.3 ~ "Moderate",
    abs_r >= 0.1 ~ "Weak",
    TRUE ~ "Negligible")
}

lag_summary <- tibble(Attribute = c("Dam Count", "Mean Dam Age", "Mean Surface Area"),
  Correlation = round(c(cor_count, cor_age, cor_size), 3),) |>
  mutate(Strength = classify_moran(Correlation))

lag_summary |> 
  knitr::kable( col.names = c("Attribute", "Spatial Lag Correlation", "Strength"),
    caption   = "Spatial Lag Correlations for Dam Attributes") |> 
    kableExtra::kable_styling("striped", full_width = TRUE, font_size = 14)

#The data indicates that the spatial lag correlation is negligible for the mean surface area but is has a strong correlation for dam count and mean dam age. This indicates that spatial neighbors in the dam count data have similar attributes to one another.
```

### Step 5.3

**Global Question:** Do neighboring tiles resemble each other overall,
or is the pattern weak?

Create a single **Moran's I scatterplot** for one dam attribute from
Step 5.2 (for example, mean year completed or dam count). Use
standardized values on both axes: the attribute on the x-axis and the
spatial lag on the y-axis. The plot should show whether points
concentrate along the diagonal or remain widely dispersed.

Your interpretation should focus on the **overall pattern**, not on
specific regions. In particular, explain whether the correlation from
Step 5.2 suggests strong, moderate, or weak spatial dependence and
whether the scatterplot supports that conclusion.

```{r}
assign_lisa_quadrants <- function(attribute_vector, spatial_lag_vector) {
  # Standardize both
  z_attr <- as.vector(scale(attribute_vector))
  z_lag  <- as.vector(scale(spatial_lag_vector))
  
  # Assign quadrants
  quadrant <- case_when(
    z_attr > 0 & z_lag > 0 ~ "HH",
    z_attr < 0 & z_lag < 0 ~ "LL",
    z_attr > 0 & z_lag < 0 ~ "HL",
    z_attr < 0 & z_lag > 0 ~ "LH",
    TRUE ~ NA_character_
  )
  
  return(list( z_attr = z_attr, z_lag = z_lag, quadrant = quadrant))
}

lag_result  <- lag_result_n   # or lag_result_year / lag_result_area
lisa_result <- assign_lisa_quadrants(lag_result$imputed, lag_result$lag)

moran_data <- data.frame(
  z     = lisa_result$z_attr,
  lag_z = lisa_result$z_lag,
  quad  = lisa_result$quadrant
)

plot_lisa_map <- function(geometry_sf, title = "LISA Map") {

  lisa_colors <- c("HH" = "#d73027", "LL" = "#4575b4",  "HL" = "#fee090", "LH" = "#91bfdb")
  
  ggplot(geometry_sf) +
    geom_sf(aes(fill = lisa), color = NA) +
    scale_fill_manual(values = lisa_colors, na.value = "gray90") +
    labs(title = title) +
    theme_void() +
    theme(legend.position = "bottom")
}
```

```{r}
assign_lisa_quadrants <- function(attribute_vector, spatial_lag_vector) {
  # Standardize both
  z_attr <- as.vector(scale(attribute_vector))
  z_lag  <- as.vector(scale(spatial_lag_vector))
  
  # Assign quadrants
  quadrant <- case_when(
    z_attr > 0 & z_lag > 0 ~ "HH",
    z_attr < 0 & z_lag < 0 ~ "LL",
    z_attr > 0 & z_lag < 0 ~ "HL",
    z_attr < 0 & z_lag > 0 ~ "LH",
    TRUE ~ NA_character_
  )
  return(list( z_attr = z_attr, z_lag = z_lag, quadrant = quadrant))
}

point_in_polygon = function(points, polygon, var){
  st_join(polygon, points) |> 
    st_drop_geometry() |> 
    count(get(var)) |> 
    setNames(c(var, "n")) |>
    left_join(polygon, by = var) |> 
    st_as_sf() 
}

tess_counts <- point_in_polygon(dams2, hexagon_conus, "ID")

lisa_result_county <- assign_lisa_quadrants(counties_pip$n, lag_county$lag)
counties_pip$lisa <- lisa_result_county$quadrant

lag_n <- compute_spatial_lag(dam_aggregate$n, W_std)
lisa_count <- assign_lisa_quadrants(lag_n$imputed, lag_n$lag)
lisa_age   <- assign_lisa_quadrants(lag_result_year$imputed, lag_result_year$lag)
lisa_size  <- assign_lisa_quadrants(lag_result_area$imputed, lag_result_area$lag)

moran_data <- data.frame(
  z = lisa_count$z_attr,
  lag_z = lisa_count$z_lag,
  quad  = lisa_count$quadrant
)

ggplot(moran_data, aes(x = z, y = lag_z, color = quad)) +
  geom_point(size = 2, alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  scale_color_manual(values = c("HH" = "#d73027", "LL" = "#4575b4",
                                "HL" = "#fee090", "LH" = "#91bfdb"),
                     na.value = "gray80") +
  labs(title = "Moran Scatterplot",
       x = "Standardized attribute", y = "Standardized spatial lag",
       color = "Quadrant") +
  theme_minimal()
```

**Global Interpretation of Spatial Patterns**

::: callout-tip
## Writing Guide: Interpreting Spatial Correlation

Use this structure to organize your response:

1.  **State your findings clearly** — cite specific correlation values
    from your Step 5.2 table. Which attributes show the strongest
    clustering?
2.  **Connect to structure** — what does the scatterplot suggest about
    broad spatial organization, without moving into local mapping yet?
3.  **Explain why** — relate to dam history, geology, hydrology, or
    policy at the regional scale
4.  **Analyze patterns** — describe what the Moran scatterplot and
    correlation values reveal about overall dam distributions
:::

Write 2–3 paragraphs (400–600 words) addressing:

**a. Spatial Clustering Patterns (Quantitative + Geographic)**

Which attributes show the strongest spatial lag correlations (from your
Step 5.2 table)? Are correlations \> 0.5 (strong) or 0.2–0.5 (moderate)?

-   **Dam *count***: Often strongly correlated because high dam density
    naturally creates neighbors with many dams.
-   **Dam *age***: Likely shows clustering if dams were built in
    regionally-coherent eras (e.g., 1930s TVA construction, 1960s–70s
    Western expansion).
-   **Dam *size***: May be clustered if geology/hydrology favors larger
    dams in specific regions.

For each attribute, describe whether the clustering seems broad or weak.
Does the correlation make geographic sense? Why should dam
characteristics be similar in neighboring areas?

**b. Moran Scatterplot Interpretation (Visualization)**

From your Moran scatterplot, describe the spatial distribution of
points:

-   Do points cluster near the diagonal, suggesting positive spatial
    autocorrelation?
-   Are there many outliers away from the diagonal, suggesting weaker
    dependence or heterogeneous neighborhoods?
-   Does the pattern appear stronger for some attributes than others?

**c. Spatial Structure Implications**

What do these correlation values and Moran scatterplot patterns tell us
about dam infrastructure spatial organization? Given your findings,
write a brief conclusion synthesizing: - Are neighboring tiles genuinely
similar in dam characteristics, or is the clustering weak? - What
management or hydrologic factors drive these patterns? - How do findings
from your Moran analysis here compare to what you observed in Q4 (dam
purposes geography)?

```{r}
#The correlation data indicates that the dam count data shows the strongest correlation. The spatial correlation for that data is "strong", with a 0.84 spatial correlation value. The mean dam age and mean surface area values returned negligible strength due to issues with n/a values. The indication of a strong spatial lag correlation for dam count can be seen on the map developed in 5.1 which shows that there is a natural collection of dams concentrated around the Mississippi. This signal strenth is not see in the mean dam age map since there is no strong pattern showing many dams created in one area at the same time. Mean surface area also does not indicate strong signals which means that dam size is variable across the country. 
#The moran scaatterplot shows a denser concentration of points in the low standardized spatial lag - low standardized attribute (LL) group which shows that there is likely a tight negative correlation with that group. This is also mostly true for the high standardized attribute - low standardized spatial lag (HL) group. The low standardized attribute - high standardized spatial lag (LH) group has more variability among points and indicates, but indicates a strong correlation. Finally, the high standardized attribute - high standardized spatial lag (HH) group has the most variability, indicating a positive corrlation that is weaker than the other groups.
#These results indicate correlation between dam count and neighboring tiles which is seen in the correlation values and in the 5.1 map. There are many dams built within a similar area, clustered around the mississippi and a few smaller hot spots focused around major streams. The results for the mean age and mean surface area are much weaker, which is not entirely surprising since dam infrastructure is more focused in one area. I was suspecting that there would be a tighter correlation for mean age in some areas since I would assume certian dam projects for one area would be funded tangetially. Overall, these findings are similar to those noted in Q4 but provide a much clearer picutre for how tighly these characteristics are correlated.
```

# Question 6: Local Spatial Autocorrelation (LISA) — Mapping Hotspots and Coldspots

**In this question, you'll analyze LISA using your recommended
tessellation from Q3 AND counties as a required comparison unit.** This
contrast (regular tessellation vs administrative boundaries) reveals
MAUP much more clearly than comparing two regular grids.

While global statistics like Moran's I (Q5) test for clustering "on
average," spatial patterns are **rarely uniform**. Some regions have
dense hotspots (e.g., TVA region with concentrated dam infrastructure),
while others are sparse coldspots (remote areas, deserts). **Local
Indicator of Spatial Autocorrelation (LISA)** decomposes global
clustering into local patterns, revealing **where** hotspots and
coldspots occur—critical for targeted management interventions.

In this question, you'll compute LISA using your **cropped recommended
tessellation** (from Q1.4) and **CONUS counties** as the alternative
support. This gives a direct test of how local clustering changes when
analysis units shift from equal-area tiles to administrative units.

::: callout-note
## LISA Concept (Week 3-1 Lectures)

LISA compares each tile's **standardized dam count** ($z_i$) to the
**standardized spatial lag** ($\text{lag}_z$):

$$\text{lag}_z = \frac{1}{w_i} \sum_{j \in N(i)} w_{ij} z_j$$

where neighbors $j \in N(i)$ are adjacent tiles, and $w_{ij}$ =
row-standardized adjacency weight (1 if touching, 0 otherwise).

The resulting **LISA quadrants** reveal: - **High-High (HH)**: High dam
count, high-count neighbors → hotspot - **Low-Low (LL)**: Low dam count,
low-count neighbors → coldspot\
- **High-Low (HL)**: High count, low neighbors → potentially isolated
cluster edge - **Low-High (LH)**: Low count, high neighbors → potential
anomaly or outlier
:::

Crop your tessellations to ensure consistent boundaries:

**a.** Use your recommended tessellation from Q3 (e.g., hexagon grid).
Clip it to the CONUS boundary with `st_filter(tessellation, boundary)`.

**b.** Use **counties** as the required alternative comparison unit (no
extra clipping needed if your counties object is already CONUS-only).

**c.** For each clipped tessellation, recompute dam counts using
point-in-polygon joining to ensure correct aggregation within the
cropped geometry.

```{r}
tess_counts <- point_in_polygon(dams2, hexagon_conus, "ID")

county_counts <- point_in_polygon(dams2, us_counties, "fip_code")
```

Compute neighbors and spatial lag for your **recommended tessellation**:

**a.** Create the binary neighbor adjacency matrix using
`st_touches(..., sparse = FALSE)`. Include only tiles within the cropped
boundary.

```{r}
W_neighbor <- standardize_weights(hexagon_conus)
```

**b.** Remove self-neighbors: `diag(W) <- 0`

**c.** Row-standardize weights: divide each row by its sum, handling
isolated tiles.

**d.** Compute spatial lag:
$\text{lag} = W_{\text{std}} \times \text{dam\_counts}$

```{r}
lag_n <- compute_spatial_lag(dam_aggregate$n, W_neighbor)
```

**e.** Standardize both dam counts and spatial lag (z-score
normalization).

```{r}
z_counts <- as.vector(scale(lag_n$imputed))
z_lag <- as.vector(scale(lag_n$lag))
```

::: callout-important
## Validation Checkpoint (Optional but Recommended)

After computing the neighbor matrix, verify it's valid:

```{r}
#| eval: false
# Check 1: Is the matrix symmetric? (for undirected adjacency)
all(W_neighbor == t(W_neighbor))

# Check 2: Are all rows row-standardized (divide by row sum)?
rowSums(W_neighbor)  # Should all be 1 (or 0 for isolated tiles)

# Check 3: Are there isolated tiles (no neighbors)?
sum(rowSums(W_neighbor) == 0)

# If any isolated tiles, document how you'll handle them in spatial lag calculation
```
:::

**LOCAL Question: WHERE specifically are high-high, low-low, high-low,
and low-high clusters?** (Answer: Geographic map showing which regions
belong to which quadrant)

**Two Visualizations:**

**a) Moran Scatterplot (Reference):** - Same as Q5, showing which tile
belongs to which quadrant - Primarily for reference; the interesting
insight is now *geographic*

**b) LISA Geographic Map (PRIMARY OUTPUT):** - Choropleth map of your
tessellation colored by LISA quadrant - HH (red) = hotspots with hotspot
neighbors → TVA region, Colorado Basin, etc. - LL (blue) = coldspots
with coldspot neighbors → Great Basin, deserts, etc. - HL (yellow) =
anomalies: high counts surrounded by sparse tiles (cluster edges) - LH
(cyan) = anomalies: sparse tiles surrounded by high-count neighbors
(rare)

**Key Interpretation:** The LISA map reveals *which specific regions*
exhibit local clustering. This is actionable: dense regions need
intensive management; sparse regions need different strategies. The
quadrant distribution (% of tiles in HH vs. LL) reveals whether the
landscape is dominated by hotspots, coldspots, or mixed.

```{r}
lag_result_tess  <- compute_spatial_lag(tess_counts$n, W_std)
lisa_result_tess <- assign_lisa_quadrants(lag_result_tess$imputed, lag_result_tess$lag)

tess_counts$z      <- lisa_result_tess$z_attr
tess_counts$lag_z  <- lisa_result_tess$z_lag
tess_counts$lisa   <- lisa_result_tess$quadrant

# (a) Moran scatterplot

plot_lisa_map <- function(geometry_sf, title = "LISA Map") {

  lisa_colors <- c("HH" = "red", "LL" = "blue",  "HL" = "yellow", "LH" = "cyan")
  
  ggplot(geometry_sf) +
    geom_sf(aes(fill = lisa), color = NA) +
    scale_fill_manual(values = lisa_colors, na.value = "gray90") +
    labs(title = title) +
    theme_void() +
    theme(legend.position = "bottom")
}

p_scatter_tess <- ggplot(tess_counts, aes(x = z, y = lag_z, color = lisa)) +
  geom_point(size = 2, alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  scale_color_manual(values = c("HH" = "#d73027", "LL" = "#4575b4",
                                "HL" = "#fee090", "LH" = "#91bfdb"),
                     na.value = "gray80") +
  labs(title = "LISA Scatter (Tessellation)",
       x = "Standardized count", y = "Standardized lag", color = "Quadrant") +
  theme_minimal()

# (b) LISA geographic map
p_map_tess <- plot_lisa_map(tess_counts, "LISA Map (Tessellation)")

p_scatter_tess + p_map_tess
```

### Step 6.4

**Comparison with required county baseline**:

**a.** Repeat Steps 6.2–6.3 for counties.

**b. REQUIRED DELIVERABLE:** Create side-by-side LISA maps of **your
chosen tessellation (from Q3)** and **counties** in the same figure. Use
identical quadrant labels and color mapping in both panels so visual
differences reflect unit choice, not styling.

**c.** **Quantitative comparison**: For both units, count the number of
polygons in each quadrant. Example output: - Hexagon: HH=15 tiles, LL=23
tiles, HL=5 tiles, LH=2 tiles - Counties: HH=205 counties, LL=1300
counties, HL=240 counties, LH=360 counties - **Interpretation**: Are
quadrant distributions similar (robust LISA pattern) or different (MAUP
sensitive at local scale)?

**d.** Interpret: Which unit reveals clustering more clearly for your
purpose? Do patterns align between your chosen tessellation and
counties, or does MAUP (modifiable areal unit problem) create divergent
local structure?

```{r}
lag_result_county  <- compute_spatial_lag(county_counts$n, W_std_county)
lisa_result_county <- assign_lisa_quadrants(lag_result_county$imputed,
                                            lag_result_county$lag)

county_counts$z     <- lisa_result_county$z_attr
county_counts$lag_z <- lisa_result_county$z_lag
county_counts$lisa  <- lisa_result_county$quadrant

p_tess   <- plot_lisa_map(tess_counts,   "LISA Map (Hexagon Tessellation)")
p_county <- plot_lisa_map(county_counts, "LISA Map (Counties)")

p_tess + p_county

list(tessellation = tess_counts |> 
       st_drop_geometry() |> 
       count(lisa),
  counties = county_counts |> 
    st_drop_geometry() |> 
    count(lisa))

#There are a higher number of tiles for each group in the counties data which is understandable since the county geometry covers larger areas in less populated spacees. There will generally be more counties in a more densely populated area which is also more likely to have increased infrastructure. The hexagon tessellation is still preferred over the county tesselation since each tile accounts for the same area of the US.
```

### Step 6.5

**Geographic Interpretation and Management Implications**

Write a 3–4 paragraph synthesis addressing:

**a.** **Geographic hotspots and coldspots**: Identify major clustering
regions: - Which dams are in HH hotspots? (E.g., TVA dams, Colorado
River dams, others?) - Which regions are LL coldspots? (Mountains,
deserts, sparse-development areas?) - What management or hydrologic
factors explain these patterns?

**Specific region example:** Choose one HH hotspot (e.g., Tennessee
Valley Authority region, Colorado River Basin, California Central
Valley) and describe three factors that explain why TVA dams or Colorado
River dams cluster. Use geographic knowledge or brief web research if
needed.

**b.** **Outliers and anomalies**: Are there notable HL or LH tiles?
What might explain them?

**c.** **Unit comparison**: Does analysis unit choice (hexagon vs.
county) affect LISA conclusions? How does MAUP operate at the local
scale?

**d.** **Implications for management**: How does LISA output inform dam
operations, reporting, or policy? - Where should multi-state
coordination efforts focus? (HH hotspots) - Which regions have simple
management structures? (LL coldspots) - How do your LISA findings from
Q6 compare to the dam purpose patterns you identified in Q4? (E.g., do
irrigation dams cluster the same way as all dams?) - Where do policies
need special attention to local anomalies? (HL/LH outliers)

```{r}
#On the tesseltation map, the middle of the US and throughout the east cost hold HH hotspots. These areas align with the Mississippi River, the Missouri River, the Rio Grande, and the east coast, which is an outlier as there is no major river but it is likely due to the high density of interconnected streams. Surprisingly, there are not as many HH areas in the west, which trends towards LL. This is due to the area's arid environment which does not allow for as many basins unlike those east of it (Spinti et al., 2023).The Colorado River Basin is an example of this, there are fewer dams within this region because of the arid climate and water availability which works best with fewer, larger reservoirs. This area also does not have as many streams as the areas with high clustering of dams which accounts for it's lower clustering.
#There are a few areas that appear to be outliers but most follow geographic and climate patterns. One outlier is in central Florida which appears to have a LH cluster. This is due to flood management efforts in the area along the Tampa Bypass Canal System. The canal was created following the flooding of the Hillsboro River after a hurricane which caused flood damage to the area (SFWMD 2026).
#The hexagon vs county does impact the LISA maps and creates more clustering in the county map. This potentially overestimates the values because the tiles cover more surface area compared to the hexagon tiles.
#The LISA map helps inform dam management by indicating areas where high dam clustering is occurring which could inform inter-state coordination of reporting and budgeting. Multi-state coordination efforts should focus on areas surrounding the Missippi and Missouri Rivers, based on the clustering on the maps. Regions with simpler management structures are the western US since they have fewer dams, though the regulatory framework for management is much more complicated due to water availability. The LISA map is most similar to the water supply, recreation, and fish and widlife maps from Q4 which indicates that these areas of policy need more attention since they show a higher clustering. Policy for outliers should be focused in areas of the West and in the outlier in Tampa, Florida. Since these areas show hotspots but are not surrounding other clustered tiles, it is likely they do not receive as much regulatory attention.


#Citations
#Spinti, R.A., Condon, L.E. & Zhang, J. The evolution of dam induced river fragmentation in the United States. Nat Commun 14, 3820 (2023). https://doi.org/10.1038/s41467-023-39194-x
#Tampa Bypass Canal System. Southest Florida Water Management District (SFWMD). (2026). https://www.swfwmd.state.fl.us/projects/tampa-bypass-canal-system#:~:text=The%20Tampa%20Bypass%20Canal%20System%20is%20a,until%20needed%20in%20times%20of%20high%20water.
```
